---
title: "TEDS-D Data Exploration and Cleaning"
author: "Valerie Jill Sydnor"
date: "4/26/2019"
output:
  pdf_document:
    toc: yes
    toc_depth: '5'
    fig_caption: yes
    keep_tex: yes
  html_document:
    code_folding: show
    highlight: pygments 
    toc: yes
    toc_depth: 4
    toc_float: yes
    df_print: kable
theme: paper
geometry: margin=1.9cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(randomForest, tree, ISLR, rpart, rattle, pROC, partykit, ggplot2, glmnet, leaps, dplyr, rsample, tm, gbm, SnowballC, RColorBrewer, wordcloud, ranger, BiocManager, keras, neuralnet, xtable, tidyverse, mapproj, psych, logisticPCA)
knitr::opts_chunk$set(options(xtable.comment = FALSE))
knitr::opts_chunk$set(comment = " ")
```


```{r eval=FALSE, include=FALSE}
# Read in Data:
drugdata <- read.csv("../TEDSD2006.csv", header=T, na.strings="-9")
```

```{r eval=FALSE, include=FALSE}
# Remove 24-Hour Detoxes and Ambulatory Detox:
drugdata.cleaned <- drugdata %>% filter(SERVSETD == "4" | SERVSETD == "5" | SERVSETD == "6" | SERVSETD == "7")
```


```{r,  eval=FALSE, include=FALSE}
# Remove Participants with Prior Treatment:
drugdata.cleaned <- drugdata.cleaned %>% filter(NOPRIOR == 0)
``` 


```{r,  eval=FALSE, include=FALSE}
# Remove Columns of Non-Interest, Low Interpretability or Redundancy:
drugdata.cleaned <- drugdata.cleaned %>% dplyr::select(-FREQ2, -FREQ3, -FRSTUSE2, -FRSTUSE3, -ROUTE2, -ROUTE3, -STFIPS, -SUB2, -SUB3, -CBSA, -PMSA, -DSMCRIT, -CASEID, -DISYR, -REGION, - NOPRIOR, -ARRESTS)
```


```{r,  eval=FALSE, include=FALSE}
# Look at NAs per Column:
sapply(drugdata.cleaned, function(x) sum(is.na(x)))
```


```{r,  eval=FALSE, include=FALSE}
# Remove Columns with too many NAs:
# DETNLF = 75% NAs
# PREG = 67%
# DETCRIM = 70%
# HLTHINS = 45%
# PRIMPAY = 52%
drugdata.cleaned <- drugdata.cleaned %>% dplyr::select(-DETNLF, -PREG, -DETCRIM, -PRIMPAY, -HLTHINS)
```


```{r,  eval=FALSE, include=FALSE}
# Remove Rows with NAs:
drugdata.cleaned <- na.omit(drugdata.cleaned)
sapply(drugdata.cleaned, function(x) sum(is.na(x))) #check
```


```{r,  eval=FALSE, include=FALSE}
# Look at Variable Class:
for(var in c("AGE", "GENDER","RACE","ETHNIC","MARSTAT", "EMPLOY", "EDUC","VET",
             "LIVARAG", "PRIMINC", "DIVISION", "SERVSETD", "METHUSE", 
             "PSOURCE", "SUB1", "ROUTE1", "FREQ1", "FRSTUSE1", "IDU", "ALCDRUG","PSYPROB")){
      drugdata.cleaned[,var] <- as.factor(drugdata.cleaned[,var])
}


drugdata.cleaned[ , grepl('FLG', names(drugdata.cleaned)) ] <- lapply(drugdata.cleaned[ , grepl('FLG' , names(drugdata.cleaned)) ], factor)    
sapply(drugdata.cleaned, function(x) class(x))
```


```{r,  eval=FALSE, include=FALSE}
# RECODE VARIABLE OF INTEREST:
for(num in c(2:8)){
drugdata.cleaned$REASON[drugdata.cleaned$REASON == num] <- 0
}
```

```{r,  eval=FALSE, include=FALSE}
# Look at Within-Variable Variability:
sapply(drugdata.cleaned, function(x) table(x))
```

```{r,  eval=FALSE, include=FALSE}
# Write Out Finalized Dataset:
write.csv(x = drugdata.cleaned, file="./Data/TEDS-D_DrugTreatment_Outcomes_Final.csv", sep=",", row.names = FALSE)
```

Read in Cleaned Data:
```{r, echo=F}
# Read in Cleaned Data
drug.data <- read.csv("./Data/TEDS-D_DrugTreatment_Outcomes_Final.csv", sep=",", header=T)

drug.data <- drug.data %>% rename(TREATCOMPLETE = REASON)

#Change Variable Classes
for(var in c("AGE", "GENDER","RACE","ETHNIC","MARSTAT", "EMPLOY", "EDUC","VET",
             "LIVARAG", "PRIMINC", "TREATCOMPLETE","DIVISION", "SERVSETD", "METHUSE", 
             "PSOURCE", "SUB1", "ROUTE1", "FREQ1", "FRSTUSE1", "IDU", "ALCDRUG","PSYPROB")){
      drug.data[,var] <- as.factor(drug.data[,var])
}

drug.data[ , grepl('FLG', names(drug.data)) ] <- lapply(drug.data[ , grepl('FLG' , names(drug.data)) ], factor)    
```

Sample Info:
```{r sample characteristics, eval=FALSE, include=FALSE}
#Demographics 
table(drug.data$AGE)
table(drug.data$GENDER)
table(drug.data$RACE)
table(drug.data$MARSTAT)
table(drug.data$EDUC)
table(drug.data$EMPLOY)

#Drug Use Information
table(drug.data$SUB1)
table(drug.data$FRSTUSE1)
table(drug.data$ALCDRUG)

#Treatment Information
table(drug.data$SERVSETD)
table(drug.data$DAYWAIT)
table(drug.data$LOS)
```

PCA on Admission Drugs:
```{r PCA Set Up, echo=F}
# Logistic Principal Component Analysis for Dimensionality Reduction of Drug Admissions Data
# Input: Drug Flags --> Flags indicate whether each drug was reported as in use at admission (0=no, 1=yes) for 18 drugs

# Step 1. Get drug flag data for input into logistic PCA
pca.data <- drug.data[, c(24:41)]
for(var in c(1:18)){
      pca.data[,var] <- as.numeric(pca.data[,var]) 
}
pca.data$ALCFLG[pca.data$ALCFLG == 1] <- 0
pca.data$ALCFLG[pca.data$ALCFLG == 2] <- 1
pca.data$COKEFLG[pca.data$COKEFLG == 1] <- 0
pca.data$COKEFLG[pca.data$COKEFLG == 2] <- 1
pca.data$MARFLG[pca.data$MARFLG== 1] <- 0
pca.data$MARFLG[pca.data$MARFLG == 2] <- 1
pca.data$HERFLG[pca.data$HERFLG== 1] <- 0
pca.data$HERFLG[pca.data$HERFLG == 2] <- 1
pca.data$METHFLG[pca.data$METHFLG== 1] <- 0
pca.data$METHFLG[pca.data$METHFLG == 2] <- 1
pca.data$OPSYNFLG[pca.data$OPSYNFLG== 1] <- 0
pca.data$OPSYNFLG[pca.data$OPSYNFLG == 2] <- 1
pca.data$PCPFLG[pca.data$PCPFLG== 1] <- 0
pca.data$PCPFLG[pca.data$PCPFLG == 2] <- 1
pca.data$HALLFLG[pca.data$HALLFLG == 1] <- 0
pca.data$HALLFLG[pca.data$HALLFLG == 2] <- 1
pca.data$MTHAMFLG[pca.data$MTHAMFLG== 1] <- 0
pca.data$MTHAMFLG[pca.data$MTHAMFLG == 2] <- 1
pca.data$AMPHFLG[pca.data$AMPHFLG== 1] <- 0
pca.data$AMPHFLG[pca.data$AMPHFLG == 2] <- 1
pca.data$STIMFLG[pca.data$STIMFLG== 1] <- 0
pca.data$STIMFLG[pca.data$STIMFLG == 2] <- 1
pca.data$BENZFLG[pca.data$BENZFLG== 1] <- 0
pca.data$BENZFLG[pca.data$BENZFLG == 2] <- 1
pca.data$TRNQFLG[pca.data$TRNQFLG== 1] <- 0
pca.data$TRNQFLG[pca.data$TRNQFLG == 2] <- 1
pca.data$BARBFLG[pca.data$BARBFLG== 1] <- 0
pca.data$BARBFLG[pca.data$BARBFLG == 2] <- 1
pca.data$SEDHPFLG[pca.data$SEDHPFLG== 1] <- 0
pca.data$SEDHPFLG[pca.data$SEDHPFLG == 2] <- 1
pca.data$INHFLG[pca.data$INHFLG== 1] <- 0
pca.data$INHFLG[pca.data$INHFLG == 2] <- 1
pca.data$OTCFLG[pca.data$OTCFLG== 1] <- 0
pca.data$OTCFLG[pca.data$OTCFLG == 2] <- 1
pca.data$OTHERFLG[pca.data$OTHERFLG== 1] <- 0
pca.data$OTHERFLG[pca.data$OTHERFLG == 2] <- 1
```

```{r PCA Tune, eval=FALSE, include=FALSE}
# Step 2. Tune Parameter K. Choose K=6!
logpca_model_k1_m5 = logisticPCA(pca.data, k = 1, m = 5)
logpca_model_k1_m5$prop_deviance_expl #27%

logpca_model_k2_m5 = logisticPCA(pca.data, k = 2, m = 5)
logpca_model_k2_m5$prop_deviance_expl #53%

logpca_model_k3_m5 = logisticPCA(pca.data, k = 3, m = 5)
logpca_model_k3_m5$prop_deviance_expl #70%

logpca_model_k4_m5 = logisticPCA(pca.data, k = 4, m = 5)
logpca_model_k4_m5$prop_deviance_expl #79%

logpca_model_k5_m5 = logisticPCA(pca.data, k = 5, m = 5)
logpca_model_k5_m5$prop_deviance_expl #84%

logpca_model_k6_m5 = logisticPCA(pca.data, k = 6, m = 5)
logpca_model_k6_m5$prop_deviance_expl #88%

logpca_model_k7_m5 = logisticPCA(pca.data, k = 7, m = 5)
logpca_model_k7_m5$prop_deviance_expl #90%

#Step 3. Tune Parameter M (if m not specified, best m is solved for at given K)
logpca_final_model = logisticPCA(pca.data, k = 6, m=0)
logpca_final_model$prop_deviance_expl #94%
save(logpca_final_model, file="./logpca_final_model.Rdata")
```

```{r PCA Get PCs, echo=F}
load("./logpca_final_model.Rdata")
PCs <- data.frame(logpca_final_model$PCs)
drug.data$PC1 <- PCs$X1
drug.data$PC2 <- PCs$X2
drug.data$PC3 <- PCs$X3
drug.data$PC4 <- PCs$X4
drug.data$PC5 <- PCs$X5
drug.data$PC6 <- PCs$X6
drug.data <- drug.data %>% select(-ends_with("FLG"))
```

Extract Training (N=40,000), Testing (N=20,000) and Validation (N=5,009) Datasets:
```{r, echo=F}
set.seed(256)

N <- nrow(drug.data)
index.train <- sample(N, 45000)
drugdata.train <- drug.data[index.train,]
drugdata.test <- drug.data[-index.train,]

N <- nrow(drugdata.test)
index.validation <- sample(N, 5009)
drugdata.validation <- drugdata.test[index.validation,]
drugdata.test <- drugdata.test[-index.validation,]
```

```{r, Neural Net Val, echo=F}
# Step 1. Define the Model
model <- keras_model_sequential() %>%
  layer_dense(units = 20, activation = "relu", input_shape = c(30)) %>% 
  layer_dense(units = 10, activation = "relu") %>%
   layer_dense(units = 10, activation = "relu") %>%
  layer_dense(units = 1, activation = "sigmoid")

# Step 2. Compile the Model
model %>% compile(
  optimizer = "rmsprop",
  loss = "binary_crossentropy",
  metrics = c("accuracy")
)

#Step 3. Train the Model
## Determine Epoch number using training and validation data sets

### Format Data
x_train <- as.matrix(drugdata.train[,-15])
y_train <- as.matrix(drugdata.train[,15])
x_validate <- as.matrix(drugdata.validation[,-15])
y_validate <- as.matrix(drugdata.validation[,15])
x_test <- as.matrix(drugdata.test[,-15])
y_test <- as.matrix(drugdata.test[,15])

### Train
neuralnet.tune <- model %>% fit(x_train, 
                                y_train,
                                epochs = 30, 
                                batch_size = 1000, 
                                validation_data = list(x_validate, y_validate)
                                )

```

```{r, echo=F}
# Step 4. Refit Model with Chosen Parameters


model <- keras_model_sequential() %>%
  layer_dense(units = 20, activation = "relu", input_shape = c(30)) %>% 
  layer_dense(units = 10, activation = "relu") %>%
   layer_dense(units = 10, activation = "relu") %>%
  layer_dense(units = 1, activation = "sigmoid")

model %>% compile(
  optimizer = "rmsprop",
  loss = "binary_crossentropy",
  metrics = c("accuracy")
)

x_train <- as.matrix(drugdata.train[,-15])
y_train <- as.matrix(drugdata.train[,15])

NN.allpredictors <- model %>% fit(x_train, y_train, epochs = 25, batch_size = 1000)

# Step 5. Evaluate 
## AUC and deviance
results <- model %>% evaluate(x_test, y_test)
results

## MCE
1-results$acc
```

Single Logistic Regression:
```{r}
for(var in c("AGE", "GENDER","RACE","ETHNIC","MARSTAT", "EMPLOY", "EDUC","VET",
             "LIVARAG", "PRIMINC", "TREATCOMPLETE","DIVISION", "SERVSETD", "METHUSE", 
             "PSOURCE", "SUB1", "ROUTE1", "FREQ1", "FRSTUSE1", "IDU", "ALCDRUG","PSYPROB", "LOS", "PC1", "PC2", "PC3", "PC4", "PC5", "DAYWAIT")){
  vardata <- drug.data[,var]
  model <- summary(glm(drug.data$TREATCOMPLETE ~ vardata, family = binomial))
      print(model)
}

```
# 1. Introduction 

## 1.1. Scope of the Problem
## 1.2. Project Goals

# 2. Methods

## 2.1. The Treatment Episode Data Set- Discharges (TEDS-D) Data
### Data Overview
### Data Cleaning
### Primary Outcome Measure
### Treatment Success Predictors
## 2.2. Study Sample Characteristics
## 2.3. Statistics
### Logistic Regression 
#### Least Absolute Shrinkage and Selection Operator
#### Elastic Net
### Random Forest
### Boosting Trees
### Neural Nets

# 3. Results

## 3.1 Models
### Logistic Regression 
#### Least Absolute Shrinkage and Selection Operator
#### Elastic Net
### Random Forest
### Boosting Trees
### Neural Nets

## 3.2. Model Comparisons

# 4. Conclusions and Implications

